---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

### Loading Required Libraries

```{r}
library("knitr")
library("ggplot2")
library("plyr")
```

### Loading Activity Data

```{r}
activityData = read.csv("activity.csv")
```

### Addding Corresponding Days for Each Date in the Data

```{r}
activityData$days = weekdays(as.Date(activityData$date,format = "%Y-%m-%d"))
```

## What is mean total number of steps taken per day?

### Daily Average and Median Steps

#### Daily Average Steps

```{r, results = 'asis'}
dailyAverage = ddply(activityData, .(date), summarise, steps = mean(steps))
dailyAverage$date = as.Date(dailyAverage$date,format = "%Y-%m-%d")
knitr::kable(dailyAverage)
```

#### Daily Median Steps

```{r}
dailyMedian = ddply(activityData, .(date), summarise, steps = median(steps))
dailyMedian$date = as.Date(dailyMedian$date,format = "%Y-%m-%d")
knitr::kable(dailyMedian)
```

### Plotting Daily Average Steps Histogram

```{r}
g = ggplot(dailyAverage, aes(date,steps))
g + geom_bar(stat = 'identity', position = 'dodge') +
        scale_x_date(date_breaks = "1 week", date_labels = "%b %d", date_minor_breaks = "1 day") +
        labs(x = 'Date') + 
        labs(y = 'Average Steps') +
        labs(title = 'Daily Average Steps')
```

## What is the average daily activity pattern?

### Time Series Plot of Interval and the Average Number of Steps Taken, Averaged Across all Days

```{r}
intAvg = ddply(na.omit(activityData), .(interval), summarise, steps = mean(steps))

g = ggplot(intAvg, aes(x = interval, y = steps))
g + geom_line(size = 0.7) + 
    labs(x = 'Interval') + 
    labs(y = 'Average Steps Across all Days') +
    labs(title = 'Average Steps for 5 Min. Intervals Across all Days')
```
    
### Interval, on Average Across all Days, Containing Maximum Number of Steps

```{r}
intAvg$interval[which.max(intAvg$steps)]
```

## Imputing missing values

### Total Number of Missing Values in the Dataset

```{r}
sum(is.na(activityData$steps)==TRUE)
```

### Imputing Missing Data

Imputing missing values using 5-minute interval average across all days for which data is available.

```{r}
impActData = activityData

for(i in 1:nrow(impActData)){
        
        if(is.na(impActData[i,1])){
                
                impActData[i,1] = intAvg$steps[which(intAvg$interval == impActData$interval[i])]
                
        }
        
}
```

### Daily Average and Median Steps

```{r}
dailyImpAverage = ddply(impActData, .(date), summarise, steps = mean(steps))
dailyImpAverage$date = as.Date(dailyImpAverage$date,format = "%Y-%m-%d")
knitr::kable(dailyImpAverage)
```

### Daily Median Steps

```{r}
dailyImpMedian = ddply(impActData, .(date), summarise, steps = median(steps))
dailyImpMedian$date = as.Date(dailyImpMedian$date,format = "%Y-%m-%d")
knitr::kable(dailyImpMedian)
```

### Plotting Daily Average Steps Histogram for Imputed Data

```{r}
g = ggplot(dailyAverage, aes(date,steps))
g + geom_bar(stat = 'identity', position = 'dodge') +
        scale_x_date(date_breaks = "1 week", date_labels = "%b %d", date_minor_breaks = "1 day") +
        labs(x = 'Date') + 
        labs(y = 'Average Steps') +
        labs(title = 'Daily Average Steps')
```

### Difference Between Imputed and Actual Data

Based on the tables provided in the next subsections, there is no difference between the imputed and actual data.

#### Difference in Average Values

```{r}
diffData = dailyAverage
diffData$steps = NULL

diffData$`Difference Between Imputed and Actual Data` = abs(dailyImpAverage$steps - dailyAverage$steps)
knitr::kable(na.omit(diffData))
```

#### Difference in Median Values

```{r}
diffData$date = dailyAverage$date
diffData$`Difference Between Imputed and Actual Data` = abs(dailyImpAverage$steps - dailyAverage$steps)
knitr::kable(na.omit(diffData))
```

## Are there differences in activity patterns between weekdays and weekends?

### Creating Weekday and Weekend Data

```{r}
days = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

weekdayData = impActData[impActData$days %in% days[1:5],]
weekendData = impActData[impActData$days %in% days[6:7],]

weekdayAvg = ddply(weekdayData, .(interval), summarise, steps = mean(steps))
weekdayAvg$day = "Weekday"

weekendAvg = ddply(weekendData, .(interval), summarise, steps = mean(steps))
weekendAvg$day = "Weekend"

avgDayActivityData = rbind(weekdayAvg,weekendAvg)
```

### Plotting Weekday and Weekend Panel Plots

```{r}
g = ggplot(avgDayActivityData, aes(x=interval,y=steps))
g + geom_line(size=0.7)+
    labs(x = 'Time Interval (minutes)') + 
    labs(y = 'Average Steps') +
    labs(title = 'Average Steps')+
    facet_grid(day~.)
```